Smoothing Algorithms for Variable Rate Models

### 2012-04-19 full_finaldraft_VRsmoothing.tex
Final draft.

\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{amsfonts}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
%\usepackage{eqparbox}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{color}

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}. An additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$.  %, comprising a product of density terms (one for each new changepoint) and a survivor function term.% In the particular (but common) case that no new changepoints occur within the interval, the density consists of a probability mass on the empty set, with weight $P(\tau_{K_n+1} > t_n | \tau_{K_n+1} > t_{n-1}, \tau_{K_n} )$.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
 \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}.% The concept of estimating the posterior for just the changepoint sequence with a particle filter rather than the joint posterior over changepoints and linear states is an example of Rao-Blackwellisation (see, e.g. \cite{Casella1996,Doucet2000}), so this algorithm is known as the Rao-Blackwellised variable rate particle filter (RB-VRPF) \cite{Godsill2007a,Christensen2012}.

@@

This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.% The resulting trajectories $x(t)^{(i)}$ for each particle are realisations from a piecewise-deterministic process, so this algorithm is named the piecewise-deterministic variable rate particle filter (PD-VRPF).

@@

%Filtering schemes which alter past changepoints -- whether using an SMC sampler or RM -- are computationally expensive, because many likelihood calculations must be conducted, for each observation from the time of the change onwards. In some cases, it may be simpler to just use a bootstrap filter with more particles.

@@

A filter conducts inference sequentially as new observations are introduced. The purpose of a smoother is to produce a revised estimate once all the observations have been made, using future values to improve upon the filter performance. Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened. Thus, a smoothing algorithm is expected to achieve significantly improved performance at changepoint estimation.

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This is the distribution approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity --- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved with a backward pass through the observations in a similar manner to the forward-backward method described in \cite{Godsill2004}.

@@

%A particle drawn from $p(\theta_{n}^{+}|y_{1:N})$ may be extended backwards by sampling from the conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$ and concatenating the past and future sequences. The new particle will be a sample from the target distribution, but the resulting approximation will still suffer from low path-space diversity before time $t_{n-1}$. Therefore, $\theta_{n-1}$ is marginalised by simply discarding the changepoints which occur before this time. This results in a particle drawn from $p(\theta_{n-1}^{+}|y_{1:N})$. The procedure then continues recursively. After a complete backwards pass through the observations, a single particle from the target distribution will have been generated. The process is then repeated until sufficient samples are obtained.

%At step $n$ for a given particle, the future changepoint sequence may be considered to be fixed. It remains to devise a method for drawing samples from the backwards conditional distribution, $p(\theta_{n}|\theta_{n}^{+}, y_{1:N})$.

@@

The second state term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$. This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}. Such a backwards Kalman filter uses the following recursions. Derivations may be found in the aforesaid references.
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \label{eq:backward_kf_predict_start} \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict_stop} \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \label{eq:backward_kf_update_start} \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
 \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}
\end{IEEEeqnarray}

Exact methods for initialising this recursion are discussed in \cite{Kitagawa1994}. However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} = m_N^{(i)}$ and $\tilde{P}_N^{(i)} = P_N^{(i)}$.

@@

The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}. Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability.      % (IS could be used, but this would require resampling steps, which would lead to a loss of path-space diversity)

@@

%The transition density term will include a delta function with unit probability mass at the point for which the past and future state trajectories match up.

%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\tilde{\theta}_{n}^{+}|\tilde{\theta}_{n}') \propto p(u_{K_n+1}|\tau_{K_n+1}, \tau_{K_n}, u_{K_n})} \nonumber \\ \times p(\tau_{K_n+1}|\tau_{K_n}, u_{K_n}) \delta_{f(x_{K_n}, v_{K_n}, \tau_{K_n}, \tau_{K_n+1})}(x_{K_n+1})
%\end{IEEEeqnarray}

@@

Each Markov chain may be initialised with the output from the previous stage. This value is itself a particle from the target distribution (albeit one from an approximation with poor diversity), meaning that no burn-in period is required.

@@

The smoothers and the final step of the filters each produce a particle approximation to the changepoint sequence posterior, $p(\theta|y_{1:N})$. Calculating a single estimator from these is no simple task. Each particle consists of a list of changepoint times and parameters, and there is no obvious way to take an average over the set to produce a single estimate. Furthermore, if a best estimate could be calculated, it is not trivial to define a distance metric between the estimated and true sequences to assess relative accuracy.

@@

In the following examples, we compare the state estimates from the VRPF, the ``filter-smoother'' and the VRPS. Filter-smoother estimates use the particle approximation from the final step of the VRPF for the changepoint distribution. In the linear-Gaussian case, an RTS smoother is used to find the state posterior distribution for each particle, $p(x_n|\theta^{(i)}, y_{1:N})$, from which state estimates are derived.

@@

where $\lambda$ introduces a mean regression effect on the trend and $\mathbf{\beta}(t)$ is standard Brownian motion (with unit diffusion constant). The jump term, $d\mathbf{J}(t)$, is zero everywhere except where jumps occur.
%
\begin{IEEEeqnarray}{rCl}
 d\mathbf{J}(t) & = & \begin{cases} \mathbf{J}_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
 \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})
\end{IEEEeqnarray}

Jumps occur at a random set of times, $\{\tau_k\}$, and are one of two types: value jumps, indicated by $u_k = 1$, and trend jumps, indicated by $u_k=2$.

@@

This model may be discretised at the observation times using standard methods (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
%
%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
%
\begin{IEEEeqnarray}{rCl}
 \mathbf{x}_n &=& A \mathbf{x}_{n-1} + \mathbf{w}_n \\
 y_n          &=& C \mathbf{x}_{n} + v_n
\end{IEEEeqnarray}

where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively.% The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
%
%\begin{IEEEeqnarray}{rCl}
% A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
% C     &=& \begin{bmatrix}1 & 0\end{bmatrix}
%\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
% Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
%                        Q_{D}             & \text{otherwise} \end{cases} \\
% Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
% q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
% q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
% q_3   &=& 1-e^{(-2 \lambda \Delta t)} \\
% R     &=& [\sigma_y^2]
%\end{IEEEeqnarray}

The times between changepoints were assumed to be exponentially distributed, with equal probability of value and trend jumps.
%
%\begin{IEEEeqnarray}{rCl}
% \IEEEeqnarraymulticol{3}{l}{p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) = P(u_k) p(\tau_k|\tau_{k-1})} \\
% p(\tau_k|\tau_{k-1})               &=& \begin{cases}\frac{1}{\alpha} \exp ( \alpha (\tau_k-\tau_{k-1}) ) & \tau_k>\tau_{k-1} \\
%                                                     0 & \tau_k < \tau_{k-1} \end{cases} \\
% P(u_k)                             &=& \begin{cases}0.5 & u_k = 1 \\ 0.5 & u_k = 2\end{cases}
%\end{IEEEeqnarray}

The algorithms were tested on artificial data simulated from this model. Observations were made on a regular time grid with a spacing of $0.0017s$. The decay constant for the distribution of inter-changepoint times is 20, and the trend mean-reversion coefficient is 5. The diffusion constant is 0.05 and the observation noise standard deviation 0.001. The jump standard deviations are 0.005 and 0.05 respectively for value and trend.

@@

The filter used $N_P = 100$ particles, and the smoother resampled $N_S = 100$ sequences. Bootstrap proposals were used for the filter. An example realisation simulated from the model is shown in figure~\ref{fig:example_data}. The trend estimates produced by the filter-smoother and VRPS are shown in figure~\ref{fig:example_state} (the equivalent graphs for value estimates are not particularly informative, as this quantity is observed).

@@

%\appendices
%\section{Initialisation of Backwards Kalman Filter} \label{app:init_backwards_KF}
%
%The backwards Kalman filter recursion may be initialised exactly by inversion of the observation model. By stacking up the last $L+1$ observations into a single vector, it is possible to write down a joint Gaussian distribution conditional on the $(N-L)$th state. This may be inverted to produce an improper Gaussian density for $x_{N-L}$,
%%
%\begin{IEEEeqnarray}{rCl}
% \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \\
% & \propto & \mathcal{N}(x_{N-L} | (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} H_{N-L}^T \Gamma_{N-L}, (H_{N-L}^T \Gamma_{N-L} H_{N-L})^{-1} )      ,
%\end{IEEEeqnarray}
%
%where $H_{N-L}$ and $\Gamma_{N-L}$ may be constructed from the system models and,
%%
%\begin{IEEEeqnarray}{rCl}
%\mathbf{y}_{N-L:N} & = & \begin{bmatrix}y_N \\ y_{N-1} \\ \vdots \\ y_{N-L} \end{bmatrix}     .
%\end{IEEEeqnarray}
%
%This should effected for the smallest value of $L$ such that the matrix inversions possible.

%\newpage
\bibliographystyle{IEEEtran}
\bibliography{D:/pb404/Dropbox/PhD/Cleanbib}
\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Bunch.tex}
\input{D:/pb404/Dropbox/PhD/Biographies/IEEE-Godsill.tex}
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\usepackage{cite}
\usepackage[pdftex]{graphicx}
\usepackage[cmex10]{amsmath}
\usepackage{amsfonts}
\usepackage{algorithmic}
\usepackage{array}
\usepackage{mdwmath}
\usepackage{mdwtab}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage{fixltx2e}
\usepackage{url}
\usepackage{color}

@@

The transition term, $p(\theta_{n \setminus n-1} | \theta_{n-1})$, has a similar form to the changepoint prior of (\ref{eq:cp_sequence_prior}) \cite{Jacobsen2006}. An additional survivor function is included to account for the condition that changepoints cannot occur before $t_{n-1}$.
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1}) = S(\tau_{K_n}, t_n) / S(\tau_{K_{n-1}+1}, t_{n-1}) } \nonumber \\
 \qquad  \qquad & & \times \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1})  \label{eq:cp_sequence_trandens}
\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
%\IEEEeqnarraymulticol{3}{l}{p(\theta_{n \setminus n-1} | \theta_{n-1})} \nonumber \\
%    & = & p(\tau_{K_{n-1}+1:K_n}, u_{K_{n-1}+1:K_n}, \tau_{K_n+1}>t_n|\tau_{K_{n-1}+1}>t_{n-1}, \tau_{K_{n-1}}, u_{K_{n-1}}) \nonumber \\
%    & = & \begin{cases} S(\tau_{K_n}, t_n) \prod\limits_{j:t_{n-1} \leq \tau_j < t_n} p(\tau_j, u_j| \tau_{j-1}, u_{j-1}, \tau_j>t_{n-1}) & K_n > K_{n-1} \\
%                        S(\tau_{K_n}, t_n) / S(\tau_{K_n}, t_{n-1}) & K_n = K_{n-1} \end{cases} \IEEEeqnarraynumspace \label{eq:cp_sequence_trandens}
%\end{IEEEeqnarray}

@@

This completes the requirements for the particle filter, resulting in the final algorithm shown in Fig.~\ref{alg:RBVRPF}.

The VRPF applied to a conditionally linear-Gaussian model estimates a posterior distribution over the changepoints only. The contribution of the linear state is calculated analytically using the Kalman recursions. This is an example of Rao-Blackwellisation (see, e.g. \cite{Casella1996,Doucet2000}).

@@

This leads to a particle filter for variable rate models with conditionally deterministic dynamics, summarised in Fig.~\ref{alg:VRPF}.

@@

@@

A filter conducts inference sequentially as new observations are introduced. The purpose of a smoother is to produce a revised estimate once all the observations have been made, using future values to improve upon the filter performance. Estimating changepoints online is a challenging task because the presence of a change may not be obvious until after it has happened. Thus, a smoothing algorithm is expected to achieve improved performance at changepoint estimation.

The target distribution for a variable rate smoothing algorithm is the posterior distribution over the entire changepoint sequence, $p(\theta|y_{1:N})$. This distribution is approximated by the final step of the VRPF. However, in the same manner as the fixed rate filter-smoother of \cite{Kitagawa1996}, this approximation is likely to lack path-space diversity --- because of the necessary resampling step in the filtering algorithm, the particles all share the same set of changepoints before a particular time, with variation only appearing for changepoints closer to $T$. For a better characterisation of the smoothing distribution, it is necessary to rejuvenate the set of particles. This is achieved with a backward pass through the observations in a similar manner to the forward-backward method described in \cite{Godsill2004}.

@@

@@

The second state term in (\ref{eq:rb-vrps_back_cond_weight}) is the likelihood, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$. This may be considered to be an improper density over $x_n$, and may be calculated analytically using a backwards Kalman filter, in a similar manner to that used in the two-filter smoother \cite{Fraser1969,Kitagawa1994,Sarkka2012,Gelb1974}. Such a backwards Kalman filter uses the following recursions, which are derived in appendix~\ref{app:backwards_KF}.
%
\begin{IEEEeqnarray}{rCl}
 p(y_{n+1:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \\
 p(y_{n:N}|x_n, \theta_{n}^+) & \propto & \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \label{eq:backward_kf_predict_start} \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \label{eq:backward_kf_predict_stop} \\
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \label{eq:backward_kf_update_start} \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
 \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \label{eq:backward_kf_update_stop}
\end{IEEEeqnarray}

Exact methods for initialising this recursion are discussed in appendix~\ref{app:init_backwards_KF}, and also in \cite{Kitagawa1994}. However, it is often sufficient to use an approximation, for example $\tilde{m}_N^{(i)} = m_N^{(i)}$ and $\tilde{P}_N^{(i)} = P_N^{(i)}$.

@@

The augmented target distribution may be sampled using Markov chain Monte Carlo (MCMC) \cite{Gilks1996}, in a similar manner to the fixed rate smoother of \cite{Bunch2012}. Metropolis-Hastings (MH) steps are conducted to draw samples, $\{\tilde{\theta}_{n}, u_{K_n}'\}$, from the target distribution by sampling the proposal and accepting the new values with a given probability. Note that IS could be used instead of MCMC, but this would require resampling steps, which would lead to a loss of path-space diversity similar to that suffered by the particle filter.

@@

@@

Each Markov chain may be initialised with the output from the previous stage. This value is itself a particle from the target distribution (albeit one from an approximation with poor path-space diversity), meaning that no burn-in period is required.

@@

The smoothers and the final step of the filters each produce a particle approximation to the changepoint sequence posterior, $p(\theta|y_{1:N})$. Calculating a single estimator from these is no simple task. Each particle consists of a list of changepoint times and parameters, and there is no obvious way to take an average over the set to produce a single estimate. Furthermore, if a best estimate could be calculated (e.g. by finding the maximum a posteriori sequence), it is not trivial to define a distance metric between the estimated and true sequences to assess relative accuracy.

@@

In the following examples, we compare the state estimates from the ``filter-smoother'' and the VRPS. Filter-smoother estimates use the particle approximation from the final step of the VRPF for the changepoint distribution. In the linear-Gaussian case, an RTS smoother is used to find the state posterior distribution for each particle, $p(x_n|\theta^{(i)}, y_{1:N})$, from which state estimates are derived.

@@

where $\lambda$ introduces a mean regression effect on the trend and $\mathbf{\beta}(t)$ is standard Brownian motion (with unit diffusion constant). The jump term, $d\mathbf{J}(t)$, is zero everywhere except when jumps occur.
%
\begin{IEEEeqnarray}{rCl}
 d\mathbf{J}(t) & = & \begin{cases} \mathbf{J}_k & t \in \{\tau_k\} \\ 0 & \text{elsewhere} \end{cases} \\
 \mathbf{J}_k  & \sim & \mathcal{N}(\mathbf{J}_k| \mathbf{0}, Q_{J,u_k})
\end{IEEEeqnarray}

$\{\tau_k\}$ is the set of random jump times, each of which is one of two types: a value jump, indicated by $u_k = 1$, or a trend jump, indicated by $u_k=2$.

@@

This model may be discretised at the observation times using standard methods, detailed in appendix~\ref{app:lg_model_discretisation}, leading to,% (see \cite{Godsill2007a,Christensen2012,Grewal2002}), leading to,
%
%Assuming observations of value only and Gaussian observation noise with standard deviation $\sigma_y^2$, the resulting discrete time dynamics are described by the following equations:% (see appendix \ref{app:model_discretisation}):
%
\begin{IEEEeqnarray}{rCl}
 \mathbf{x}_n &=& A \mathbf{x}_{n-1} + \mathbf{w}_n \\
 y_n          &=& C \mathbf{x}_{n} + v_n
\end{IEEEeqnarray}

where the $\mathbf{w}_n$ and $v_n$ are Gaussian random variables with covariance matrixes $Q_n$ and $R$ respectively. $Q_n$ is a function of the sequence of jump times and their types, giving rise to the desired conditionally linear-Gaussian structure. % The time between observations times is denotes $\Delta t = t_{n} - t_{n-1}$.
%
%\begin{IEEEeqnarray}{rCl}
% A     &=& \begin{bmatrix}1 & \frac{1}{\lambda}(1-e^{(-\lambda \Delta t)} \\ 0 & e^{(-\lambda \Delta t)}\end{bmatrix} \\
% C     &=& \begin{bmatrix}1 & 0\end{bmatrix}
%\end{IEEEeqnarray}
%\begin{IEEEeqnarray}{rCl}
% Q_n   &=& \begin{cases}Q_{D} + Q_{J,u_k} & \exists k : \tau_k \in [t_{n-1},t_n]\\
%                        Q_{D}             & \text{otherwise} \end{cases} \\
% Q_{D} &=& \frac{\sigma^2}{2 \lambda}\begin{bmatrix} q_1 & q_2 \\ q_2 & q_3\end{bmatrix} \\
% q_1   &=& \frac{1}{\lambda^2}(2 \lambda \Delta t - (3 - e^{(-\lambda \Delta t)})(1 - e^{(-\lambda \Delta t)})~) \\
% q_2   &=& \frac{1}{\lambda} (1-e^{(-\lambda \Delta t)})^2 \\
% q_3   &=& 1-e^{(-2 \lambda \Delta t)} \\
% R     &=& [\sigma_y^2]
%\end{IEEEeqnarray}

The times between changepoints are assumed to be exponentially distributed, with equal probability of value and trend jumps.
%
%\begin{IEEEeqnarray}{rCl}
% \IEEEeqnarraymulticol{3}{l}{p(\tau_k, u_k|\tau_{k-1}, u_{k-1}) = P(u_k) p(\tau_k|\tau_{k-1})} \\
% p(\tau_k|\tau_{k-1})               &=& \begin{cases}\frac{1}{\alpha} \exp ( \alpha (\tau_k-\tau_{k-1}) ) & \tau_k>\tau_{k-1} \\
%                                                     0 & \tau_k < \tau_{k-1} \end{cases} \\
% P(u_k)                             &=& \begin{cases}0.5 & u_k = 1 \\ 0.5 & u_k = 2\end{cases}
%\end{IEEEeqnarray}

The algorithms were tested on artificial data simulated from this model. Observations were made on a regular time grid with a spacing of $0.0017s$. The decay constant for the distribution of inter-changepoint times is 20, and the trend mean-reversion coefficient is 5. The diffusion constant is 0.05 and the observation noise standard deviation 0.001. The jump standard deviations are 0.005 and 0.05 respectively for value and trend. (Terms defined in appendix~\ref{app:lg_model_discretisation})

@@

The filter used $N_F = 100$ particles, and the smoother resampled $N_S = 100$ sequences. Bootstrap proposals were used for the filter. An example realisation simulated from the model is shown in figure~\ref{fig:example_data}. The trend estimates produced by the filter-smoother and VRPS are shown in figure~\ref{fig:example_state} (the equivalent graphs for value estimates are not particularly informative, as this quantity is observed).

@@

\appendices
\section{The Backwards Kalman Filter} \label{app:backwards_KF}

A backwards Kalman filter \cite{Fraser1969,Kitagawa1994} is used to calculate the joint likelihood term, $p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)$, when the state dynamics are linear-Gaussian. Assume that at time $t_{n+1}$ there is a Gaussian density such that,
%
\begin{equation}
 p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+) = \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1})  .
\end{equation}

The state transition model is linear-Gaussian, so the backwards predictive density is given by,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+)} \nonumber \\
  & = & \int p(y_{n+1:N}|x_{n+1}, \tilde{\theta}_{n}^+)  p(x_{n+1}|x_n, \tilde{\theta}_{n}^+) dx_{n+1} \nonumber \\
  & = & \int \tilde{Z}_{n+1} \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
  &   & \qquad \mathcal{N}(x_{n+1}|A_{n+1} x_{n}, Q_{n+1}) dx_{n+1} \nonumber \\
  & = & \int \tilde{Z}_{n+1} ||A_{n+1}^{-1}|| \mathcal{N}(x_{n+1}|\tilde{m}_{n+1}, \tilde{P}_{n+1}) \nonumber \\
  &   & \qquad \mathcal{N}(x_{n}|A_{n+1}^{-1} x_{n+1}, A_{n+1}^{-1} Q_{n+1} A_{n+1}^{-T}) dx_{n+1} \nonumber \\
  & = & \tilde{Z}_{n}^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-)
\end{IEEEeqnarray}

where,
%
\begin{IEEEeqnarray}{rCl}
 \tilde{m}_n^- & = & A_{n+1}^{-1} \tilde{m}_{n+1} \\
 \tilde{P}_n^- & = & A_{n+1}^{-1} (\tilde{P}_{n+1} + Q_{n+1}) A_{n+1}^{-T} \\
 \tilde{Z}_n^- & = & \tilde{Z}_{n+1} ||A_{n+1}||^{-1}   .
\end{IEEEeqnarray}

$||A||$ is used to denote the magnitude of the determinant of $A$, and it is assumed that $A$ is non-singular.

The update step simply involves multiplying by the likelihood of the observation at time $t_n$,
%
\begin{IEEEeqnarray}{rCl}
\IEEEeqnarraymulticol{3}{l}{p(y_{n:N}|x_n, \tilde{\theta}_{n}^+)} \\
 & = & p(y_{n+1:N}|x_n, \tilde{\theta}_{n}^+) p(y_{n}|x_n, \tilde{\theta}_{n}^+) \nonumber \\
 & = & \tilde{Z}_n^- \mathcal{N}(x_n|\tilde{m}_n^-, \tilde{P}_n^-) \mathcal{N}(y_n|C_n x_n, R_n) \nonumber \\
 & = & \tilde{Z}_n \mathcal{N}(x_n|\tilde{m}_n, \tilde{P}_n)
\end{IEEEeqnarray}

where,
%
\begin{IEEEeqnarray}{rCl}
 \tilde{\mu}_n & = & C_n \tilde{m}_n^- \\
 \tilde{S}_n   & = & C_n \tilde{P}_n^- C_n^T + R_n \\
 \tilde{K}_n   & = & \tilde{P}_n^- C_n^T \tilde{S}_n^{-1} \\
 \tilde{m}_n   & = & \tilde{m}_n^- + \tilde{K}_n (y_n - \tilde{\mu}_n) \\
 \tilde{P}_n   & = & \tilde{P}_n^- - \tilde{K}_n \tilde{S}_n \tilde{K}_n^T \\
 \tilde{Z}_n   & = & \tilde{Z}_n^- \mathcal{N}(y_n|\tilde{\mu}_n, \tilde{S}_n)   .
\end{IEEEeqnarray}

All steps use only standard Gaussian identities.

\section{Initialisation of the Backwards Kalman Filter} \label{app:init_backwards_KF}

The backwards Kalman filter recursion may be initialised exactly by inversion of the observation model using the method of \cite{Kitagawa1994}. By stacking up the last $L+1$ observations into a single vector, it is possible to write down a joint Gaussian distribution conditional on the $(N-L)$th state. This may be inverted to produce an improper Gaussian density for $x_{N-L}$,
%
%\begin{IEEEeqnarray}{rCl}
% \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
% & = & \sqrt{\frac{|(H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1}|}{|\Gamma_{N-L}|}} \nonumber \\
% & & \times \mathcal{N}(x_{N-L} | (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N}, \nonumber \\
% & & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} )      ,
%\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 \IEEEeqnarraymulticol{3}{l}{p(y_{N-L:N}|x_{N-L}) = \mathcal{N}( \mathbf{y}_{N-L:N} | H_{N-L} x_{N-L}, \Gamma_{N-L} ) } \nonumber \\
 & = & \sqrt{\frac{|V_{N-L}|}{|\Gamma_{N-L}|}} \mathcal{N}(x_{N-L} | \nu_{N-L}, V_{N-L} )
\end{IEEEeqnarray}
\begin{IEEEeqnarray}{rCl}
 V_{N-L} & = & (H_{N-L}^T \Gamma_{N-L}^{-1} H_{N-L})^{-1} \\
 \nu_{N-L} & = & V_{N-L} H_{N-L}^T \Gamma_{N-L}^{-1} \mathbf{y}_{N-L:N}     ,
\end{IEEEeqnarray}
